\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-12 challenge.tex
partial recursor

Family polymorphism enables modular reuse via late binding:
code of a base family can be reused by a derived family,
because what fields referenced by that code unfold to is polymorphic to
the enclosing family.

This flexibility, however, prevents \emph{definitional equality}
that one takes for granted when programming in a proof assistant.
In \cref{fig:stlc-nonmechanized}, a proof assistant supporting family
polymorphism cannot equate references to the substitution function and
its definition (i.e., a pattern match against four cases),
as a derived family may modify the definition of the function by adding
new cases.
Without the ability to unfold the substitution function, how can the
programmer even prove the substitution lemma?
%
The problem is compounded by the occasional need to override fields in derived families,
which also precludes the unfolding of definitions.

@@

Interactive theorem proving and tactic programming are central to
the idiomatic programming experience in a proof assistant.
A language design integrating family polymorphism should be compatible
with these programming styles.
In particular, it should be possible in our tool to
incrementally navigate through vernacular commands and, moreover, construct proofs
with common tactics while getting instant feedback on the proof state,
even in the middle of a family definition.
In addition, it should remain possible to perform computation in the extended Gallina language.

>>>>

Family polymorphism enables modular reuse via late binding:
code of a base family can be reused by a derived family,
because fields referenced by that code have meanings polymorphic to the
enclosing family.

This flexibility, however, prevents \emph{definitional equality}
that one takes for granted when programming in a proof assistant.
In \cref{fig:stlc-nonmechanized}, a proof assistant supporting family
polymorphism cannot equate references to the substitution function and
its definition (i.e., a pattern match against four cases),
as a derived family may modify the definition of the function by adding
new cases.
Without the ability to unfold the substitution function, how can the
programmer even prove the substitution lemma?
%
The problem is compounded by the occasional need in derived families to override fields,
which is potentially at odds with being able to use equalities over the fields.

@@

Interactive theorem proving and tactic programming are central to
the idiomatic programming experience in a proof assistant.
A language design integrating family polymorphism should be compatible
with these programming styles.
In particular, it should be possible in our tool to
incrementally navigate through vernacular commands and, moreover, construct proofs
with common tactics while getting instant feedback on the proof state,
even in the middle of a family definition.
In addition, it should remain possible to normalize terms that mention families.

### 2022-10-12 lang-design.tex
partial recursor

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLC_fix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.
For example, executing the last line, \lsti{Check STLC_fix.typesafe}, will
display the type-safety theorem of the fixpoints extension.

@@

In family "STLC_fix", inductive type "tm" \emph{further binds} the "tm" type
in family "STLC". It is generated by five constructors: four inherited from
"STLC" and a fifth called "tm_fix".

Crucially, the types of the constructors are polymorphic to their enclosing family---%
their meaning is \emph{late bound}.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lsti[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lsti{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "STLC_fix" include those constructed from "tm_fix".

@@

Ordinarily, an inductive type is not extensible: it is exhaustively
generated by its constructors and has no more inhabitants beyond those
they construct.
This idea is captured by the eliminator (aka recursor) associated with
an inductive type.
For example, if "tm" were defined as an ordinary inductive type, then
its eliminator would have the following type:

\begingroup
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall (P : tm -> Type), [@\hlcase{tm-recursor-case-var}@](\forall x, P (tm_var x))[@\tikzmarkend{tm-recursor-case-var}@] -> [@\hlcase{tm-recursor-case-unit}@]P tm_unit[@\tikzmarkend{tm-recursor-case-unit}@] -> [@\hlcase{tm-recursor-case-abs}@](\forall x t, P t -> P (tm_abs x t))[@\tikzmarkend{tm-recursor-case-abs}@] ->
  [@\hlcase{tm-recursor-case-app}@](\forall t1 t2, P t1 -> P t2 -> P (tm_app t1 t2))[@\tikzmarkend{tm-recursor-case-app}@] -> \forall t, P t
\end{lstlisting}
\endgroup

\noindent
The eliminator would allow for function definitions by recursion
and proofs by induction, over "tm", that exhaustively
handle the four cases corresponding to each constructor.
The (dependent) return type~"P" is called the \emph{motive} of the recursion.
%Exhaustivity of induction would follow, trivially.

With inductive types made extensible, exhaustivity of induction becomes
in jeopardy, however.
To reconcile the tension without requiring redefinition or rechecking of
case handlers (challenge \ref{chg:extensible-v-exhaustive}),
our design introduces the \lsti{FRecursion} and \lsti{FInduction} commands.
The key idea is to allow case handlers to be added retroactively, should
inductive types be extended, and to allow recursion and induction (which are defined in terms of case handlers) to be late bound.

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC" is further bound by the "subst" in "STLC_fix":
the four cases from "STLC" are automatically inherited and reused,
with "STLC_fix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by induction on "hasty" and because "STLC_fix" adds
a constructor to "hasty", the programmer is required to handle this extra case
in the proof of "substlem" in family "STLC_fix".

@@

OO inheritance allows the late binding of method names.
Family polymorphism generalizes the power of OO inheritance by allowing the
late binding of all names nested within families, including those referring to types.
We have seen that late binding of "tm" allows the "tm" constructors in "STLC"
to be reused in "STLC_fix" to construct terms containing "tm_fix".

As another example, consider the type of "st_beta" in family "STLC".
It refers to "subst", whose meaning is late bound.
When "st_beta" is inherited into family "STLC_fix", "st_beta" has a type that now
refers to the "subst" function in "STLC_fix", where "subst" is defined
by handling all the five cases known to that family.
Thus, late binding of "subst" allows the derived family to reuse "st_beta"
to $\beta$-reduce abstractions whose body is a fixpoint.

Importantly, a name is late bound only within a family that defines or extends it.
Outside such families, the name can be accessed only by explicitly
specifying a family that contains it.
For example, the last line of \cref{fig:stlc-mechanized} accesses "typesafe"
with "STLC_fix.typesafe".
Executing the command prints

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLC_fix.typesafe: \forall t t' T, STLC_fix.steps t t' -> STLC_fix.hasty STLC_fix.empty t T ->
                   STLC_fix.value t' \/ \exists t'', STLC_fix.step t' t''.
\end{lstlisting}

\noindent
In this type, all references to nested names are qualified by "STLC_fix", as desired.

@@

However, with \lsti{subst} being late bound, \lsti{subst} can no longer
be unfolded (challenge~\ref{chg:latebind-v-defeq}):
the definition of \lsti{subst}, as a recursive function,
varies across families, yet a derived family should be able to reuse the
proof of the \lsti{ht_unit} case
even when it has to modify the definition of \lsti{subst}.
Without the ability to unfold \lsti{subst},
how can the programmer make progress in this proof, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{subst tm_unit x t' = tm_unit}, as a proposition
(\lsti{Prop}), can still be proved.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Based on this insight, our plugin automatically generates and proves
propositional equalities for recursive functions defined with \lsti{FRecursion}.
The plugin also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the generated
propositional equalities.

Note that the definitional equality on \lsti{subst} is available
outside those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound.
In contrast, outside those families, \lsti{subst} can be referenced only
by specifying a family that contains it,
such as \lsti{STLC.subst} or \lsti{STLC_fix.subst}.
So as far as Coq's type checker is concerned, \lsti{tm_unit} and
\lsti{STLC_fix.subst tm_unit x t'} are the same thing---Coq equates them
definitionally by unfolding \lsti{STLC_fix.subst} and performing
$\beta$-reduction.

@@

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, the programmer may want to override
the proof of a lemma with a different proof strategy than it was proved
in the base family.
%
Overriding is potentially in conflict with having equalities on late bound names.

@@

Our plugin does allow the overriding of transparent definitions that are
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking requires unfolding that field
should be overridden too.
We give an example in a supplemental appendix.

@@

In \cref{fig:stlc-mechanized}, just because the "typesafe" theorem in
"STLC" is inherited and reused by "STLC_fix", it does not follow that
in "STLC_fix" the programmer can use "typesafe" to prove "progress".
If the programmer did, then they would be committing the logical fallacy
of circular reasoning: the proof of "progress" would depend on "typesafe",
yet the proof of "typesafe" depends on "progress".
%
Such circularity would easily lead to logical inconsistency:

\begin{centered}
\begin{minipage}{36ex}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
Family A.
FLemma f : False. Proof. Admitted.
FLemma g : False. Proof. apply f. Qed.
End A.
\end{lstlisting}
\end{minipage}
\hspace{3.5em}
\begin{minipage}{36ex}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
Family B extends A.
FLemma f : False. Proof. apply g. Qed.
End A.
[@~@]
\end{lstlisting}
\end{minipage}
\end{centered}

@@

To ensure the soundness of logical reasoning, the type system requires
that in a derived family, the context in which a field is defined be
preserved from the base family.
In "STLC", "progress" is in the context of "typesafe".
Per the requirement, this relationship must be preserved into
"STLC_fix", which prevents the proof of "progress" from depending on
"typesafe" in "STLC_fix".

@@

Families can be readily reused to construct larger extensions that
\emph{mix in}~\cite{mixin-1990} the functionalities of the individual
families.
In \cref{fig:stlc-isorec-prod}, family "STLC_fix_isorec" is declared as
an "STLC" extension that mixes in "STLC_fix" and "STLC_isorec".
As a result, \lsti{STLC_fix_isorec.typesafe} readily gives the
type-safety theorem of STLC equipped with fixpoints and iso-recursive types.

@@

In the presence of extensible inductive types, mixin composition may
also create an obligation to retrofit the mixins with new case handlers.
In \cref{fig:stlc-isorec-prod}, family "STLC_prod_isorec" is composed
of two mixins: "STLC_prod", which extends the inductive type "ty" with a
new constructor "ty_prod", and "STLC_isorec", which introduces a
function "tysubst" recursively defined on "ty".
Hence, for exhaustivity, it is required that a composition of
"STLC_prod" and "STLC_isorec" should additionally handle the "ty_prod"
case in "tysubst".

\noindentparagraph{Tactics for eliminating inductive types.}

>>>>

\cref{fig:stlc-mechanized} shows how STLC and its extension with
fixpoints can be mechanized using our plugin, in a style envisioned in
\cref{fig:stlc-nonmechanized}.
The base family "STLC" hosts the STLC metatheories, from abstract syntax
to the type-safety theorem.
Family "STLCFix", derived from "STLC", makes adjustments as needed by a fixpoints extension:
it adds new constructors to the inductive types (\lsti{FInductive}) and
adds new cases to the recursive functions (\lsti{FRecursion}) and
induction proofs (\lsti{FInduction}).
Existing constructors and cases, as well as those definitions and
theorems that need no adjustments ("ty", "env", "empty", "steps", and
"typesafe"), are automatically inherited and reused.
In particular, executing the last command, \lsti{Check STLCFix.typesafe},
displays the type-safety theorem of the fixpoints extension.

@@

In family "STLCFix", inductive type "tm" \emph{further binds} the "tm" type
in family "STLC". It is generated by five constructors: four inherited from
"STLC" and a fifth called "tm_fix".

Crucially, the types of the constructors are polymorphic to their enclosing family:
the types mention "tm", and the meaning of "tm" is \emph{late bound}.
For example, although "tm_app" is defined in the base family,
it can be used to construct applications of fixpoints in the derived family:
%such as \lsti[mathescape]{tm_app (tm_fix "f" (tm_abs "x" $\dadada$)) tm_unit},
its type \lsti{tm -> tm -> tm} says "tm_app" can be applied to any
Gallina terms of type "tm",
which in family "STLCFix" include those constructed from "tm_fix".

@@

Ordinarily, an inductive type is not extensible: it is exhaustively
generated by its constructors and has no more inhabitants beyond those
they construct.
This idea is captured by the eliminator (aka \emph{recursor})
associated with an inductive type.
For example, if "tm" were defined as an ordinary inductive type, then
its eliminator would have the following type:

\begin{centered}
\begin{minipage}{.71\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
tm_rec : \forall (P : tm -> Set), [@\hlcase{tm-recursor-case-var}@](\forall x, P (tm_var x))[@\tikzmarkend{tm-recursor-case-var}@] -> [@\hlcase{tm-recursor-case-unit}@]P tm_unit[@\tikzmarkend{tm-recursor-case-unit}@] ->
         [@\hlcase{tm-recursor-case-abs}@](\forall x t, P t -> P (tm_abs x t))[@\tikzmarkend{tm-recursor-case-abs}@] ->
         [@\hlcase{tm-recursor-case-app}@](\forall t1, P t1 -> \forall t2, P t2 -> P (tm_app t1 t2))[@\tikzmarkend{tm-recursor-case-app}@] -> \forall t, P t
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
The eliminator would enable function definitions by recursion
and proofs by induction, over "tm", that exhaustively
handle the four cases corresponding to each constructor.
The (dependent) return type~"P" is called the \emph{motive} of the recursion.
%Exhaustivity of induction would follow, trivially.

With inductive types made extensible, exhaustivity of induction becomes
in jeopardy, however.
In particular, recursor "tm_rec" should no longer be allowed, because
its type mentions "tm", whose meaning is late bound, yet
"tm_rec" purports to claim \lsti{\forall t, P t} given only four case handlers.

To reconcile the tension without requiring redefinition or rechecking of
case handlers~(\ref{chg:extensible-v-exhaustive}),
our design introduces the \lsti{FRecursion} and \lsti{FInduction} commands.
The key idea is to allow case handlers to be added retroactively, should
inductive types be extended, and to allow recursion and induction (which are defined in terms of case handlers) to be late bound.

As an example, consider the substitution function "subst", defined using
\lsti!FRecursion!.
The \lsti{on} clause specifies that recursion is over "tm".
The \lsti{motive} clause suggests that the recursive function being
defined has type \lsti{tm -> id -> tm -> tm}.
%
The "subst" function in "STLC" is further bound by the "subst" in "STLCFix":
the four cases from "STLC" are automatically inherited and reused,
with "STLCFix" retroactively adding a fifth case to form the new "subst" function.
%
For exhaustivity, it is a static error if the programmer fails to
further bind "subst" and define this fifth case.
The type system does this check by examining if the inductive type "tm",
over which "subst" is recursively defined, is further bound in the same family.

@@

\noindent
The programmer can use tactic programming to discharge the goal.
%The idiomatic programming experience with Coq is largely preserved.
Because the lemma is by induction on "hasty" and because "STLCFix" adds
a constructor to "hasty", the programmer is required to handle this extra case
in the proof of "substlem" in family "STLCFix".

@@

OO inheritance allows the late binding of method names.
Family polymorphism generalizes the power of OO inheritance by allowing the
late binding of all names nested within families, including those referring to types.
We have seen that late binding of "tm" allows the "tm" constructors in "STLC"
to be reused in "STLCFix" to construct terms containing "tm_fix".

As another example, consider the type of "st_beta" in family "STLC".
It refers to "subst", whose meaning is late bound.
When "st_beta" is inherited into family "STLCFix", "st_beta" has a type that now
refers to the "subst" function in "STLCFix", where "subst" is defined
by handling all the five cases known to that family.
Thus, late binding of "subst" allows the derived family to reuse "st_beta"
to $\beta$-reduce abstractions whose body is a fixpoint.

Importantly, a name is late bound only within a family that defines or extends it.
Outside such families, the name can be accessed only by explicitly
specifying a family that contains it.
For example, the last line of \cref{fig:stlc-mechanized} accesses "typesafe"
with "STLCFix.typesafe".
Executing the command prints

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLCFix.typesafe: \forall t t' T, STLCFix.steps t t' -> STLCFix.hasty STLCFix.empty t T ->
                   STLCFix.value t' \/ \exists t'', STLCFix.step t' t''.
\end{lstlisting}

\noindent
In this type, all references to nested names are qualified by "STLCFix", as desired.

@@

However, with \lsti{subst} being late bound, \lsti{subst} can no longer
be unfolded~(\ref{chg:latebind-v-defeq}):
the definition of \lsti{subst}, as a recursive function,
varies across families, yet a derived family should be able to reuse the
proof of the \lsti{ht_unit} case
even when it has to modify the definition of \lsti{subst}.
Without the ability to unfold \lsti{subst},
how can the programmer make progress in this proof, then?

A key observation is that although late binding
prevents \emph{definitional equality} on \lsti{subst}, it does not
affect \emph{propositional equality}.
That is, \lsti{subst tm_unit x t' = tm_unit}, as a proposition
(\lsti{Prop}), should still hold.
After all, how \lsti{subst} is defined on \lsti{tm_unit} does not vary
from a base family to a derived family.
%
Based on this insight, our plugin automatically generates
propositional equalities for each recursive function defined with \lsti{FRecursion},
making them available as axioms for use by the rest of the current family.
The plugin also provides a tactic \lsti{fsimpl} that enables, for instance,
simplifying \lsti{subst tm_unit x t'} to \lsti{tm_unit}.
Under the hood, \lsti{fsimpl} works by rewriting using the postulated
propositional equalities.

Note that the definitional equality on \lsti{subst} is available
outside those families that contain "subst".
Within those families, the meaning of \lsti{subst} is late bound, so only
propositional equality is available.
In contrast, outside those families, \lsti{subst} is always referenced
by specifying a family that contains it,
such as \lsti{STLC.subst} or \lsti{STLCFix.subst}.
As far as the type checker is concerned, \lsti{tm_unit} and
\lsti{STLCFix.subst tm_unit x t'} are the same thing---the type checker
equates them definitionally by unfolding \lsti{STLCFix.subst} and
performing normalization.

@@

In an OO language, a subclass can override methods of a superclass.
Similar expressiveness is useful for metatheory mechanization, too.
%
For example, in a derived family, the programmer may want to override
the proof of a lemma with a different proof strategy than it was proved
in the base family.
%
Overriding is potentially incompatible with having equalities on late bound names.

@@

Our plugin does allow the overriding of transparent definitions that are
explicitly marked as \lsti{Overridable} by the programmer.
Overriding is made safe by requiring that when an overridable field is
overridden, code whose type checking involves unfolding that field
should be overridden too.
We expect this feature to be used only occasionally.
We give an example in \cref{sec:coqexample}.

@@

In \cref{fig:stlc-mechanized}, just because the "typesafe" theorem in
"STLC" is inherited and reused by "STLCFix", it does not follow that
in "STLCFix" the programmer can use "typesafe" to prove "progress".
If the programmer did, then they would be committing the logical fallacy
of circular reasoning: the proof of "progress" would depend on "typesafe",
yet the proof of "typesafe" depends on "progress".
%
Such circularity would easily lead to logical inconsistency:

\begin{centered}
\definecolor{codecomment-color}{HTML}{556B2F}
\begin{minipage}{36ex}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
Family A.[@\tanc{A-inconsistency-start}@]
FLemma f : False. Proof. Admitted.
FLemma g : False. Proof. apply f. Qed.
End A.[@\tanc{A-inconsistency-end}@]
\end{lstlisting}
\end{minipage}
\makeline[22pt]{A-inconsistency-start}{A-inconsistency-end}[codecomment-color!50]
\hspace{3.5em}
\definecolor{codecomment-color}{HTML}{FF00FF}
\begin{minipage}{36ex}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
Family B extends A.[@\tanc{B-inconsistency-start}@]
FLemma f : False. Proof. apply g. Qed.
End B.[@\tanc{B-inconsistency-end}@]
[@~@]
\end{lstlisting}
\end{minipage}
\makeline[.5\textwidth+23pt]{B-inconsistency-start}{B-inconsistency-end}[codecomment-color!50]
\end{centered}

@@

To ensure the soundness of logical reasoning (\ref{chg:self-v-consistency}),
the type system requires that in a derived family, the context in which
a field is defined be preserved from the base family.
In "STLC", "progress" is in the context of "typesafe".
Per the requirement, this relationship must be preserved into
"STLCFix", which prevents the proof of "progress" from depending on
"typesafe" in "STLCFix".

@@

Families can be readily reused to construct larger extensions that
\emph{mix in}~\cite{mixin-1990} the functionalities of the individual
families.
In \cref{fig:stlc-isorec-prod}, family "STLCFixIsorec" is declared as
an "STLC" extension that mixes in "STLCFix" and "STLCIsorec".
As a result, \lsti{STLCFixIsorec.typesafe} is readily the proof of the type-safety
theorem of an STLC equipped with fixpoints and iso-recursive types.

@@

In the presence of extensible inductive types, mixin composition may
also create an obligation to retrofit the mixins with new case handlers.
In \cref{fig:stlc-isorec-prod}, family "STLCProdIsorec" is composed
of two mixins: "STLCProd", which extends the inductive type "ty" with a
new constructor "ty_prod", and "STLCIsorec", which introduces a
function "tysubst" recursively defined on "ty".
Hence, for exhaustivity, it is required that a composition of
"STLCProd" and "STLCIsorec" should additionally handle the "ty_prod"
case in "tysubst".

\noindentparagraph{Tactics support for constructors.}

Coq provides tactics for proving injectivity and disjointness of constructors
(i.e., \lsti{injection} and \lsti{discriminate}).
The proof terms generated by the tactics involve exhaustively
matching on the constructors of an inductive type, so
they do not work for extensible inductive types (\ref{chg:extensible-v-exhaustive}).
%
In principle, the programmer could use \lsti{FInduction} to prove
injectivity and disjointness.
But that would be tedious, and the programmer would have to revisit
the induction proof every time the inductive type is extended.
%
Having to use \lsti{FInduction} is unsatisfying, also because
injectivity and disjointness ought to hold
regardless of whether inductive types are extensible.

To provide a streamlined programming experience (\ref{chg:tooling}),
our plugin offers two tactics, \lsti{finjection} and \lsti{fdiscriminate}.
For example, in a proof state that contains a manifestly false assumption
\lsti{H : tm_var "x" = tm_abs "y" t}, the programmer can use
\lsti{fdiscriminate H} to obtain \lsti{False} and thus discharge
the current goal, just as they would with \lsti{discriminate H} if "tm"
were not extensible.

\noindentparagraph{Partial recursors.}

The tactics \lsti{finjection} and \lsti{fdiscriminate} are powered by
\emph{partial recursors} that our plugin automatically generates for all
inductive types defined with \lsti{FInductive}.

As analyzed earlier, ordinary recursors, such as "tm_rec", are impossible
within a family in which the name of the inductive type is late bound.
However,
a key observation is that extensible inductive types still admit a
weakened form of recursors where the motive is an \lsti{option} type.
For example, within family "STLC", the partial recursor for
"tm" has the following type:

\begin{centered}
\begin{minipage}{.96\textwidth}
\newcommand\hlcase[1]{\tikzmarkin[disable rounded corners=true,set fill color=yellow!18,set border color=yellow!18]{#1}(0.00,-0.05)(-0.00,0.20)}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLC_tm_prec : \forall (P : tm -> Set), [@\hlcase{tm-partialrec-case-var}@](\forall x, option (P (tm_var x)))[@\tikzmarkend{tm-partialrec-case-var}@] -> [@\hlcase{tm-partialrec-case-unit}@]option (P tm_unit)[@\tikzmarkend{tm-partialrec-case-unit}@] ->
               [@\hlcase{tm-partialrec-case-abs}@](\forall x t, option (P t) -> option (P (tm_abs x t)))[@\tikzmarkend{tm-partialrec-case-abs}@] ->
               [@\hlcase{tm-partialrec-case-app}@](\forall t1, option (P t1) -> \forall t2, option (P t2) -> option (P (tm_app t1 t2)))[@\tikzmarkend{tm-partialrec-case-app}@] ->
               \forall t, option (P t)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
As expected, "STLC_tm_prec" is axiomatized along with
four (propositional) equalities, one for each constructor of "tm".
For instance, the equality for the "tm_abs" constructor is as follows:

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall x t P H1 H2 H3 H4,
STLC_tm_prec P H1 H2 H3 H4 (tm_abs x t) = H3 x t (STLC_tm_prec P H1 H2 H3 H4 t)
\end{lstlisting}

%\noindent
Importantly, the partial recursor is compatible with the late binding of "tm" in its type.
When "STLC_tm_prec" is inherited into family "STLCFix", all the previous four equalities
still hold, and a trivial, fifth equality is made available:

\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
\forall x t P H1 H2 H3 H4, STLC_tm_prec P H1 H2 H3 H4 (tm_fix x t) = None
\end{lstlisting}

Although partial recursors appear weaker than ordinary recursors,
there is power in restraint:
they offer a principled way to derive injectivity and disjointness of
constructors, while allowing extensibility.
Partial recursors allow for injective mappings from values of a late bound inductive
type, like "tm", to values of an ordinary inductive type, like \lsti{\nat},
the injectivity and disjointness of whose constructors are readily
available.
%
For example, in a proof state with the assumption \lsti{H : tm_var "x" = tm_abs "y" t},
running \lsti{fdiscriminate H} first uses \lsti{H} to obtain the equality

\begin{centered}
\begin{minipage}{\textwidth}
\begin{lstlisting}[basicstyle=\fontsize{8.25}{9}\ttfamily]
STLC_tm_prec (\lambda[@\,@]_, \nat) (\lambda[@\,@]_, Some 1) (Some 2) (\lambda[@\,@]_[@\,@]_[@\,@]_, Some 3) (\lambda[@\,@]_[@\,@]_[@\,@]_[@\,@]_, Some 4) (tm_var "x") =
STLC_tm_prec (\lambda[@\,@]_, \nat) (\lambda[@\,@]_, Some 1) (Some 2) (\lambda[@\,@]_[@\,@]_[@\,@]_, Some 3) (\lambda[@\,@]_[@\,@]_[@\,@]_[@\,@]_, Some 4) (tm_abs "y" t)
\end{lstlisting}
\end{minipage}
\end{centered}

\noindent
and then uses the axiomatized equalities about "STLC_tm_prec" to
rewrite the equality above, obtaining \lsti{Some 1 = Some 4},
from which \lsti{False} easily follows.

### 2022-10-12 stlc-isorec-prod.tex
partial recursor

\definecolor{codecomment-color}{HTML}{FF007F}

\begin{lstlisting}
Family STLC_isorec extends STLC.[@\tanc{STLCisorec-start}@]
[@\hfill\codecomment{STLC extended with iso-recursive types}@]
FInductive ty : Set +=
| ty_var : id -> ty | ty_rec : id -> ty -> ty.

@@

[@\dadada@]

End STLC_isorec.[@\tanc{STLCisorec-end}@]
\end{lstlisting}

\makeline[0pt]{STLCisorec-start}{STLCisorec-end}[codecomment-color!50]

@@

\begin{lstlisting}
Family STLC_prod extends STLC.[@\tanc{STLCprod-start}@]
[@\hfill\codecomment{STLC extended with products}@]
FInductive ty : Set +=
| ty_prod : ty -> ty -> ty.

@@

[@\dadada@]

End STLC_prod.[@\tanc{STLCprod-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCprod-start}{STLCprod-end}[codecomment-color!50]

\vspace{-5pt}

@@

\begin{lstlisting}
Family STLC_isorec_prod extends STLC[@\tanc{STLCisorecprod-start}@]
using STLC_prod, STLC_isorec.
[@\hfill\ \ \codecomment{STLC extended with iso-recursive types and products}@]
FRecursion tysubst on ty
motive \lambda(_ : ty), id -> ty -> ty.
Case ty_prod := [@\dadada@].[@\hfill\codecomment{Substitution on product types}@]
End tysubst.

End STLC_isorec_prod.[@\tanc{STLCisorecprod-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCisorecprod-start}{STLCisorecprod-end}[codecomment-color!50]

@@

>>>>

\definecolor{codecomment-color}{HTML}{DD033F}

\begin{lstlisting}
Family STLCIsorec extends STLC.[@\tanc{STLCIsorec-start}@]
[@\hfill\codecomment{STLC extended with iso-recursive types}@]
FInductive ty : Set +=
| ty_var : id -> ty | ty_rec : id -> ty -> ty.

@@

[@\dadada@][@\hfill\codecomment{Other adjustments}@]
End STLCIsorec.[@\tanc{STLCIsorec-end}@]
\end{lstlisting}

\makeline[0pt]{STLCIsorec-start}{STLCIsorec-end}[codecomment-color!50]

\vspace{-12pt}

\definecolor{codecomment-color}{HTML}{008B8B}

\begin{lstlisting}
Family STLCFixIsorec extends STLC[@\tanc{STLCIsorecfix-start}@]
using STLCFix, STLCIsorec.
[@\hfill\ \ \codecomment{STLC extended with fixpoints and iso-recursive types}@]
End STLCFixIsorec.[@\tanc{STLCIsorecfix-end}@]
\end{lstlisting}

\makeline[0pt]{STLCIsorecfix-start}{STLCIsorecfix-end}[codecomment-color!50]

@@

\begin{lstlisting}
Family STLCProd extends STLC.[@\tanc{STLCProd-start}@]
[@\hfill\codecomment{STLC extended with products}@]
FInductive ty : Set +=
| ty_prod : ty -> ty -> ty.

@@

[@\dadada@][@\hfill\codecomment{Other adjustments}@]
End STLCProd.[@\tanc{STLCProd-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCProd-start}{STLCProd-end}[codecomment-color!50]

\vspace{-10pt}

@@

\begin{lstlisting}
Family STLCProdIsorec extends STLC[@\tanc{STLCIsorecProd-start}@]
using STLCProd, STLCIsorec.
[@\hfill\ \ \codecomment{STLC extended with products and iso-recursive types}@]
FRecursion tysubst on ty
motive \lambda(_ : ty), id -> ty -> ty.
Case ty_prod := [@\dadada@].[@\hfill\codecomment{Substitution on product types}@]
End tysubst.

End STLCProdIsorec.[@\tanc{STLCIsorecProd-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCIsorecProd-start}{STLCIsorecProd-end}[codecomment-color!50]

\vspace{-10pt}

\definecolor{codecomment-color}{HTML}{708090}

\begin{lstlisting}
Family STLCFixProdIsorec extends STLC[@\tanc{STLCIsorecFixProd-start}@]
using STLCFix, STLCProdIsorec.[@\hfill\codecommentstart{STLC extended}@]
[@\hfill\codecommentend{with fixpoints, products, and iso-recursive types}@]
End STLCFixProdIsorec.[@\tanc{STLCIsorecFixProd-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCIsorecFixProd-start}{STLCIsorecFixProd-end}[codecomment-color!50]

@@

\vspace{-10pt}

### 2022-10-12 stlc-mechanized.tex
partial recursor

\begin{lstlisting}[firstnumber=77]
Family STLC_fix extends STLC.[@\tanc{STLCfix-start}@]
[@\hfill\codecomment{STLC extended with fixpoints}@]
FInductive tm : Set +=
| tm_fix : id -> tm -> tm.

@@

End STLC_fix.[@\tanc{STLCfix-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCfix-start}{STLCfix-end}[codecomment-color!50]

@@

\begin{lstlisting}[firstnumber=111]
Check STLC_fix.typesafe.[@\tanc{Check-STLCfix-typesafe}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{Check-STLCfix-typesafe}{Check-STLCfix-typesafe}[codecomment-color!35]

>>>>

\begin{lstlisting}[firstnumber=77]
Family STLCFix extends STLC.[@\tanc{STLCFix-start}@]
[@\hfill\codecomment{STLC extended with fixpoints}@]
FInductive tm : Set +=
| tm_fix : id -> tm -> tm.

@@

End STLCFix.[@\tanc{STLCFix-end}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{STLCFix-start}{STLCFix-end}[codecomment-color!50]

@@

\begin{lstlisting}[firstnumber=111]
Check STLCFix.typesafe.[@\tanc{Check-STLCFix-typesafe}@]
\end{lstlisting}

\makeline[.5\textwidth+9pt]{Check-STLCFix-typesafe}{Check-STLCFix-typesafe}[codecomment-color!35]
